Solution of Differential Algebraic Equations with Local Constraints on the Latent Space of an Autoencoder

### 2019-07-11 paper/paper.tex
Edits to the paper

@@

Our proposed approach is to build upon modern machine learning
techniques to aid in equation discovery. An autoencoder is trained on
the original material dataset to learn a lower-dimensional
representation of the underlying manifold. The model is designed to be
directly incorporated into a macroscale simulation of physical balance
laws, where now the traditional partial differential equations are being
solved on the learned latent space. This provides a more robust
numerical algorithm, while putting the onus of programming part of the
simulation itself onto the computer: tens of thousands of lines of human
written code consisting of empirical fits and branching logic are
replaced by code generated by the learned model. The methodology will
first be demonstrated on datasets based on thermodynamic intensive
variables, which are already challenging to analyze, before being
extended to richer datasets such as experimental images or molecular
dynamics. We stress that the proposed approach is only utilizing machine
learning on material data: the simulation is indeed solving physical
balance laws. Through deep learning approaches combined with
domain-specific goals, we can learn feature sets that are both
interpretable and directly apply to analyzing larger systems.

We train an autoencoder to determine the lower-dimensional
representation of the EOS surface using modern deep learning techniques
without using any phase labels. The decoder end of the model is then
plugged back into the balance laws, eliminating the constraint equation.
\end{abstract}

@@

Representing complex physical systems with multiple species, phases,
chemical reactions, and heterogenous scale effects poses an
intractability problem to modeling. How do we analyze complex
experimental data and \emph{ab initio} simulations and derive new
macroscale relations to describe the systems? How do we then implement
these relations in the rest of our analyses and production-scale
computer simulations
Accurately simulating multispecies reaction and transport with
undergoing phase transitions is a critical aspect to a variety of
problems.
This work is motivated by advancing simulation techniques for subsurface energy systems such as shale
and oceanic natural gas reservoirs in shales or engineered geothermal
systems.

@@

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e =& \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}
(In this paper, we use $\partial_t$, $\mathrm{d}/\mathrm{d}t$,
and $\dot{()}$ to denote differentiation with respect to time
interchangeably depending on context.
The complexity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

% {\bf RESERVOIR PHYSICS LIT REVIEW HERE; Other applications;
% plasticity, }
One of the simulators with the widest-encompassing physical modules is
the TOUGH family of solvers that have been developed at Lawrence
Berkeley National Lab.
The methodology is based on a primary variable switching method in 
The simulator of THAT GUY WITH DUAL VARIABLES I MET AT AGU
parameterizes a simplified hydrate system using density and enthalpy
as the primary variables.
Higher order representations of flow are also a problem to tackle
anisotropy and increase the computational efficiency of the spatial
discretization\cite{hannon_enhanced_2018}. 
Multiphase flow was done with finite elements in \cite{yang_fully_2014}.
THAT MORIDIS FEM PAPER
However, the complexity of the state representation restricts
development to a 2-phase

@@

% {\bf ML LIT REVIEW HERE} % TODO
Representation learning methods such as StyleGAN and G|FLOW have shown
the ability to find meaningful dimensions in the latent space,
e.g. the popular ability to transform age, race, and gender of human
faces in a smooth way.
But, can this directions be used meaningfully?

@@

Part of the new modern software engineering framework involves replacing
human-written code with computer-written code. We are applying this
mindset from industry problems in the Artificial
Intelligence space into scientific computing.

@@

@@

We build a test dataset off of the IAPWS solutions. The proposed
methodology applies to experimental data directly, or preexisting
compendiums of empirical fits and theoretical relations.
Note that, even if a relation can be theoretical derived, that does
not mean that the analytical relationship yields the best program
implementation: it is often the case that numerial approximations are
superior to closed-form mathematical solutions in computational
efficiency, accuracy, and numerical stability.

In section 2, the balance of mass and energy will be described for a
single multiphase material. We will discuss the current state of the
art algorithm for deriving relationships for multiphase materials, and
then the primary-variable switching algorithm that is used in
state-of-the art solvers. The equation will then be recast as a
constraint problem in th above form.
In section 3, we will demonstrate applying the autoencoder learning
methodology to solve the pendulum,
which can be considered the archetypical constraint problem.
In section 4, the autoencoder methodology wil lbe applied to the
equation of state of water. In section 5, the new simulation process
will be described, wherein the learned decoders are placed into the
balance laws, forming newly discovered equations on learned
representations. The software implementation of the training and
simulation programs are detailed in section 6. The new simulation
framework on sets of learned representations is evaluated on
multiple datasets of varying extents of water with a
hyperparameter search on a suite of test problems to prouce the discussion of section 7, as
well as a systematic automated testing procedure of a
machine-dominated programming methodology.
Active research and software development directions are
concluded in the final section.

@@

For this preliminary work, we consider only a single-beaker problem
without flow, the idealization of the continuous stirred-tank reactor (CSTR), as illustrated
in Figure \ref{fig:cstr}. Let \(\rho\) be the material density, \(T\) be the
temperature, \(p\) be the material pressure, \(h\) be the material
enthalpy, and \(u\) be the internal energy. There are two equations for
balance of mass and energy.
\begin{align}
\partial_t \rho & = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t \rho u & = \mathbf{k}_T(T_\infty-T) + s 
\end{align}
The beaker is exposed to a reservoir with \(p_\infty\) and \(T_\infty\)
via a mixed boundary condition straw with transport and conduction
coefficients \(\mathbf{k}_p\), \(\mathbf{k}_T\). Mass and heating
sources and sinks are the terms \(r\) and \(s\).\footnote{We express the
  material by enthalpy instead of internal energy; they are related by
  the relation \(\rho u = \rho h - p\). Future work may work directly
  with internal energy.}
We are neglecting upstreaming in the connection between the reservoir
and the container.

@@

We also consider the phase equilibria: beyond the scale of resolving
individual bubbles, it is possible
for a control volume to be a mixture of two phase coexisting.
(For a very fine scale simulation, e.g. at the pore scale, bubbles
would be explicitly modeled.)
We introduce a saturation $S_{phase}\in[0,1]$ for each phase;.
The material can exist on this section of the surface as a mixture of
the two phases. Thus, its density is calculated by a mixing the
properties of all phases at that point in the phase diagram:
\begin{equation}
  \rho = S_{gas} \rho_{gas} + S_{liq} \rho_{liq} + S_{ice}
    \rho_{ice} + S_{crit} \rho_{crit}
  \end{equation}
Each phase
and equilibrium surface constrains the possible saturations. For example, on the boiling equilibrium surface there are two phases with some mixture
saturations \(S_{gas}=1-S_{liq}\) and $S_{ice}=S_{crit}=0$. 
For water, three phases can exist in equilibrium. Given more species, it is possible for an arbitrary number of phases
to be in equilibrium; there is a quadruple point in methane-water
hydrates.

@@

{\bf Remark:} Many popular forms of EOS do not even have closed form
expressions. The family of cubic equations of state for nonideal
gases, such as a Peng-Robinsion equation \cite{peng_new_1976}, are
expressed as functions for $p$ of $V$ (a proxy for $\rho$) and
$T$. Since $\rho$ is rarely a primary varibale, these expressions must be inverted for
$\rho$ given $p$ by solving the cubic equation. This can be performed
analytically, as is done in TOUGH+, but often polynomial roots can
be best solved numerically. When the equations become more
complicated, such as when there are gas mixtures, the equations of
state must be solved numerically. TOUGH+ employs a Richardson
iteration \cite{moridis_users_2014-2}.
Note how this is a numerical
iteration {\em inside} of another numerical iteration that is trying
to differentiation this equation state. Approximations
abound--the mathematical theory is not straight foward. Passing
automatic differentiation objects through these computations is
difficult in practice. Equations are
often solved faster and more accurately by numerical
approximations--why not completely rewrite the equations?

@@

The typical methodology is to use the empirical relations for density
and enthalpy \(h\) as a function of pressure and temperature and solve
the DAE for \(p\) and \(T\) implicitly:
\begin{align}
\text{Solve for}\, p(t)\, \text{and}\, T(t)\, \text{such that:}\\
\partial_t \rho(p,T) & = \nabla \cdot \mathbf{k}\nabla p + r\\
\partial_t \rho u(p,T) & = \nabla \cdot \mathbf{k'}\nabla T + s
\end{align}
The complication is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.

\begin{tabular}{l|c|c}
  State & $X_1$ & $X_2$ \\
  \hline
  Ice & $p$ & $T$ \\
  Liquid & $p$ & $T$ \\
  Gas & $p$ & $T$ \\
  Supercritical & $\rho$ & $T$ \\
  Ice-Liquid & $S_{ice}$ & $T$ \\
  Ice-Gas & $S_{ice}$ & $T$ \\
  Gas-Liquid & $S_{gas}$ & $T$ \\
  Ice-Gas-Liquid & $S_{ice}$ & $S_{gas}$ 
\end{tabular}

  This requires a statemachine. For the complete states of water, it is shown in Figure \ref{fig:statemachine}.
  \begin{figure}
    \centering
    \includegraphics[width=4in]{../slides/phase_4.png}
    \caption{\label{fig:statemachine}States and possible transitions for water.}
\end{figure}
Formulate balance equations as
\begin{equation}
R(X ; phase) = 0
\end{equation}
Solve the differentiable part:
\[\frac{\partial R}{\partial X}\Delta X^{k+1/2} = -R(X^k ; phase^k)\]
and iterate on the nondifferential part:
\[phase^{k+1},X^{k+1} = statemachine(X^{k+1/2} ; phase^k)\]
The phase switching logic has the form of a switch case:

\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{switch}\NormalTok{ phase_old:}
  \ControlFlowTok{case}\NormalTok{ gas:}
    \ControlFlowTok{if}\NormalTok{ p,T crossing boundary:}
\NormalTok{      phase_new = liquid_gas}
  \ControlFlowTok{case}\NormalTok{ liquid:}
    \ControlFlowTok{if}\NormalTok{ p,T crossing boundary:}
\NormalTok{      phase_new = liquid_gas}
  \ControlFlowTok{case}\NormalTok{ liquid_gas:}
    \ControlFlowTok{if}\NormalTok{ S_gas >= }\DecValTok{1}\NormalTok{:}
\NormalTok{      phase_new = gas}
    \ControlFlowTok{if}\NormalTok{ S_liquid >= }\DecValTok{1}\NormalTok{:}
\NormalTok{      phase_new = gas}
\end{Highlighting}
\end{Shaded}

@@

We instead replace our problem with two balance laws and one unknown
constraint:
\begin{align}
\text{Solve for}\, \rho(t), \, h(t), \, p(t),\, \text{and}\, T(t)\, \text{satisfying:}\\
\partial_t \rho &  = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t (\rho h-p) & = \mathbf{k}_T(T_\infty-T)  + s + h\left(\mathbf{k}_p(p_\infty - p) + r\right) \\
\text{such that they lie on the material EOS,}\\
\rho,h,p,T & \in \{\text{EOS}\}
\end{align}
The \emph{ground truth} is the experimental data in the first place; the
branching curve fit is only one realization of representing the data. We
have demonstrated that it is not possible to represent any two of these
in terms of the other two; and even if so, it is not necessarily a good
and robust fit. What if we search for a new fit?

@@

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
x^2 + y^2 & = R^2
\end{align}
Note that we are considering the nonlinear pendulum.

There are many ways to solve this canonical equation: Lagrange
multipliers (\(f(t)\) is the multiplier in the way we have written it
above), penalties, change of variables, etc. The change of variables
method exploits the Lagrangian mechanics formulation, where we state
the Lagrangian,
\begin{equation}
L = \frac{1}{2}m\left(\dot{x}^2 + \dot{y}^2\right) - m g y
\end{equation}
with the constraint that \(x\) and \(y\) lie on the path.
We can introduce a new variable \(\theta\) that parameterizes
\(x=R\cos\theta\) and \(y=R\sin\theta\) and rederive the equations of
motion with the familiar,
\begin{align}
\text{Solve for}\, \theta(t) \ \text{satisfying:} \\
\frac{\mathrm{d}}{\mathrm{d}t} \frac{\partial L}{\partial
  \dot{\theta}} =
  \frac{\partial L}{\partial \theta}
\end{align}
In this example, this lets us collapse the problem into a
single-component second-roder differential equation.

In our methodology, we do not have an explicit equation for the
constraint, but a dataset of $X$ of $(x,y)$ pairs that are on the manifold. For
our proof-of-concept test, we manufacture this data for the circular
pendulum to verify against the analytical solution:
\begin{equation}
X = \left\{ (R\cos\theta_i, R\sin\theta_i)\; | \;\theta_i \in [-5\pi/4,\pi/4] \right\}
\end{equation}
We also try a ``goofier'' looking curve with no known closed form
solution to exercise the methodology,

(For the equation of state problem, this is analogous to the dataset
of $p,T,\rho,h$ 4-tuples that make the plot in Figure \ref{fig:pTrho}.)

@@

We'll use the autoencoder to generate the parameterization of \(x(q)\)
and \(y(q)\). To do this, we optimize the loss function,
\begin{equation}
L(x) = \left\| x-D(E(x)) \right\|
\end{equation}

When we want to solve the dynamics, we can treat the autoencoder the
same as \(x(\theta)\), plugging in
\[L(x(q),v(q,\dot{q}))\]
Taking the partial derivatives with respect to the feature variable and
its rate, the components to the equation of motion are build the DAE we know from physics:
\[\frac{\mathrm{d}}{\mathrm{d}t}\frac{\partial L}{\partial \dot{q}} = \frac{\partial L}{\partial q}\].
By the chain rule, the velocities are related to the
feature space by
\begin{equation}
\dot{x} = \frac{\mathrm{d}x(q)}{\mathrm{d}t} = \frac{\partial x}{\partial q}\frac{\mathrm{d}q}{\mathrm{d}t}
\end{equation}
and similarly for the \(y\) direction.
Then plug it into a Runge Kutta (trapezoidal) to solve with Newton's
method:
\[\frac{\partial L}{\partial \dot{q}}_i - \Delta t \frac{\partial L}{\partial q}_i = \frac{\partial L}{\partial \dot{q}}_0 + \Delta t \frac{\partial L}{\partial q}_0\]

Soltutions are shown in Figure \ref{fig:pendulum}.
\begin{figure}
\centering
\includegraphics[width=5in]{../figures/pendulum_q.png}
\caption{\label{fig:pendulum}Solution of the pendulum on an arbitrary latent space.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../slides/classifier_network.png}
\caption{\label{fig:classifyingnetwork}The architecture of the classifier network designed for
  unsupervised labeling of phases.}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=3in]{../slides/classifier_good.jpg}
  \caption{\label{fig:classified}Result of the unsupervised phase classifier. Compare this plot to the original labeled dataset Figure \ref{fig:pTrho}.}
\end{figure}
Because the identity of the phases,XXXXXXXX
For comparison purposes, we group together liquid-gas-supercritical as one contiguous label.

@@

\hypertarget{header-n3353}{%
\subsection{Pseudo-code}\label{header-n3353}}

>>>>

\usepackage{algorithm}

@@

Our proposed approach is to build upon modern machine learning
techniques to aid in equation discovery. An autoencoder is trained on
the original material dataset to learn a lower-dimensional
representation of the underlying manifold. The model is designed to be
directly incorporated into a macroscale simulation of physical balance
laws, where now the traditional partial differential equations are being
solved on the learned latent space.
We train an autoencoder to determine the lower-dimensional
representation of the EOS surface using modern deep learning techniques
without using any phase labels. The decoder end of the model is then
plugged back into the balance laws, eliminating the constraint equation.
This provides a more robust
numerical algorithm, while putting the onus of programming part of the
simulation itself onto the computer: tens of thousands of lines of human
written code consisting of empirical fits and branching logic are
replaced by code generated by the learned model. The methodology will
first be demonstrated on datasets based on thermodynamic intensive
variables, which are already challenging to analyze, before being
extended to richer datasets such as experimental images or molecular
dynamics. We stress that the proposed approach is only utilizing machine
learning on material data: the simulation is indeed solving physical
balance laws. Through deep learning approaches combined with
domain-specific goals, we can learn representations that are both
interpretable and directly apply to analyzing larger systems.
We illustrate the methodology on two traditional problems: 1) the
pendulum, and 2) the continuous stirred tank reactor.

\end{abstract}

@@

Representing complex physical systems with multiple species, phases,
chemical reactions, and heterogenous scale effects poses an
intractability problem to modeling. How do we analyze complex
experimental data and \emph{ab initio} simulations and derive new
macroscale relations to describe the systems? How do we then implement
these relations in the rest of our analyses and production-scale
computer simulations
Accurately simulating multispecies reaction and transport with
undergoing phase transitions is a critical aspect to a variety of
problems.
This work is motivated by addressing challenges in the simulation of subsurface energy systems, such as shale
and oceanic natural gas reservoirs or engineered geothermal
systems.

@@

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e =& \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve \cite{queiruga_simulation_2019}. 
(In this paper, we use $\partial_t$
and $\dot{()}$ to denote differentiation with respect to time
for compactness.)
The complexity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

% {\bf RESERVOIR PHYSICS LIT REVIEW HERE; Other applications;
% plasticity, }
One of the simulators with the widest-encompassing physical modules is
the TOUGH family of solvers that have been developed at Lawrence
Berkeley National Lab.
The methodology is based on a primary variable switching method in 
The simulator of THAT GUY WITH DUAL VARIABLES I MET AT AGU
parameterizes a simplified hydrate system using density and enthalpy
as the primary variables.
Higher order representations of flow are also a problem to tackle
anisotropy and increase the computational efficiency of the spatial
discretization\cite{hannon_enhanced_2018}. 
Multiphase flow was done with finite elements in \cite{yang_fully_2014}.
THAT MORIDIS FEM PAPER SAY MORE ABOUT IT

@@

% {\bf ML LIT REVIEW HERE} % TODO
Representation learning methods such as StyleGAN
\cite{karras_style-based_2018} and Flow/Glow \cite{kingma_glow:_2018} have shown
the ability to find meaningful dimensions in the latent space,
e.g. the popular ability to transform age, race, and gender of human
faces in a smooth way.
But, can this directions be used meaningfully?

Recent review papers have highlighted open applicaitons of machine
learning to subsurface systems \cite{bergen_machine_2019} and
climate scale systems  \cite{reichstein_deep_2019}.
History matching on the latent space of a GAN was demonstrated on a
synthetic geomorphology by Mosser, 2019 \cite{mosser_deepflow:_2019}.

Many approaches try to use machine learning to learn the dynamics of a
system.
The method of Dynamic Mode Decomposition
to learn a linear operator based on time series data \cite{tu_dynamic_2014,kutz_dynamic_2016}.

Extending this space with known features, such as partial derivatives,
\cite{ one of kutz's students}
Our presented approach differs from this approach by saying that we
know the dynamics {\em a priori}, but do not how to represent the
current state in a useful way. (Our approach will be later extended to
learn dynamics in a second step using the learned representation as a
fixed input.)

@@

Part of the new modern software engineering framework involves replacing
human-written code with computer-written code. We are applying this
mindset from industry problems in the artificial
intelligence space into scientific computing.

@@

There are infinitely many representations of a system. The simplest and clearist description is not neccessarily the best way
to solve it, and vice versa.

@@

We build a dataset using the IAPWS solutions. The proposed
methodology applies to experimental data directly, or preexisting
compendiums of empirical fits and theoretical relations.
Note that, even if a relation can be theoretical derived, that does
not mean that the analytical relationship yields the best program
implementation: it is often the case that numerial approximations are
superior to closed-form mathematical solutions in computational
efficiency, accuracy, and numerical stability.

In Section 2, the balance of mass and energy will be described for a
single multiphase material. We will discuss the current state of the
art algorithm for deriving relationships for multiphase materials, and
then the primary-variable switching algorithm that is used in
state-of-the art solvers. The equation will then be recast as a
constraint problem in th above form.
In Section 3, we will demonstrate applying the autoencoder learning
methodology to solve the pendulum,
which can be considered the archetypical constraint problem.
In Section 4, the autoencoder methodology wil lbe applied to the
equation of state of water. In Section 5, the new simulation process
will be described, wherein the learned decoders are placed into the
balance laws, forming newly discovered equations on learned
representations. The software implementation of the training and
simulation programs are detailed in Section 6. The new simulation
framework on sets of learned representations is evaluated on
multiple datasets of varying extents of water with a
hyperparameter search on a suite of test problems to prouce the discussion of Section 7, as
well as a systematic automated testing procedure of a
machine-dominated programming methodology.
Active research and software development directions are
concluded in the final section.

@@

For this preliminary work, we consider only a single-beaker problem
without flow, the idealization of the continuous stirred-tank reactor (CSTR), as illustrated
in Figure \ref{fig:cstr}. 
In the analysis of geological fluids, this idealization is used in the
assumption of well-mixed pore spaces at the macroscale.  Let \(\rho\)
be the material density, measured in \si{kg/m^3}; \(T\) be the
temperature, measured in \si{K}; \(p\) be the material pressure,
measured in \si{Pa}; \(h\) be the material
specific enthalpy, measured in \si{J/kg}; and \(u\) be the internal
energy, measured in \si{J/kg}, and related to the enthalpy by  \(\rho u = \rho h - p\). There are two equations for
balance of mass and energy.
\begin{align}
\partial_t \rho & = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t \rho u & = \mathbf{k}_T(T_\infty-T) + s + h q_{mass}
\end{align}
The beaker is exposed to a reservoir with \(p_\infty\) and \(T_\infty\)
via a mixed boundary condition straw with transport and conduction
coefficients \(\mathbf{k}_p\), \(\mathbf{k}_T\). 
We are neglecting upstreaming in the connection between the reservoir
and the container. Mass and heating
sources and sinks are the terms \(r\) and \(s\).

@@

We also consider the phase equilibria: beyond the scale of resolving
individual bubbles, it is possible
for a control volume to be a mixture of multiple coexisting
phases.
(For a very fine scale simulation, e.g. at the pore scale, bubbles
would be explicitly modeled.)
We introduce a saturation $S_{phase}\in[0,1]$ for each phase;.
The material can exist on this section of the surface as a mixture of
the two phases. Thus, its density is calculated by a mixing the
properties of all phases at that point in the phase diagram:
\begin{align}
  \rho = & S_{gas} \rho_{gas} + S_{liq} \rho_{liq} + S_{ice}
  \rho_{ice} + S_{crit} \rho_{crit} \\
  \rho h =& S_{gas} \rho_{gas} h_{gas} + S_{liq} \rho_{liq} h_{liq} + S_{ice}
  \rho_{ice} h_{ice} + S_{crit} \rho_{crit} h_{crit}
\end{align}
Each phase
and equilibrium surface constrains the possible saturations. For example, on the boiling equilibrium surface there are two phases with some mixture
saturations \(S_{gas}=1-S_{liq}\) and $S_{ice}=S_{crit}=0$. 
For water, three phases can exist in equilibrium. Given more species, it is possible for an arbitrary number of phases
to be in equilibrium; there is a quadruple point in methane-water
hydrates.

Note that the definition of enthalpy as a per-mass quanitity means the mixture needs to be weighted by the
density: $h_{mix}=\sum{\rho_a h_aS_a}/\sum{rho_aS_a}$.

@@

{\bf Remark:} Many popular forms of EOS do not even have closed form
expressions. The family of cubic equations of state for nonideal
gases, such as the Peng-Robinsion equation \cite{peng_new_1976}, are
expressed as functions for $p$ of $V$ (a proxy for $\rho$) and
$T$. Since $\rho$ is rarely a primary varibale, these expressions must be inverted for
$\rho$ given $p$ by solving the cubic equation. This can be performed
analytically, as is done in TOUGH+, but often polynomial roots can
be best solved numerically. When the equations become more
complicated, such as when there are gas mixtures, the equations of
state must be solved numerically. TOUGH+ employs a Richardson
iteration \cite{moridis_users_2014-2}.
Note how this is a numerical
iteration {\em inside} of another numerical iteration that is trying
to differentiation this equation state. Approximations
abound--the mathematical theory is not straight foward. Passing
automatic differentiation objects through these computations is
difficult in practice. Equations are
often solved faster and more accurately by numerical
approximations--why not completely rewrite the equations?

@@

The typical methodology is to use the empirical relations for density
and enthalpy \(h\) as a function of pressure and temperature and solve
the DAE for \(p\) and \(T\) implicitly:
\begin{align}
\text{Solve for}\, p(t)\, \text{and}\, T(t)\, \text{such that:}\\
\partial_t \rho(p,T) & = \nabla \cdot \mathbf{k}\nabla p + r\\
\partial_t \rho(p,T) h(p,T)-p & = \nabla \cdot \mathbf{k'}\nabla T + s
\end{align}
The complication is that the functions \(\rho(p,T)\) and \(u(p,T)\)
are not well defined functions due to the presence of phase changes that
yield sharp discontinuities in \(p,T\), as shown in Figure 1 for
water.
Each state requires a potentially different set of primary
variables. For a pure water system, this would have to be:
\begin{center}
\begin{tabular}{l|c|c}
  State & $X_1$ & $X_2$ \\
  \hline
  Ice & $p$ & $T$ \\
  Liquid & $p$ & $T$ \\
  Gas & $p$ & $T$ \\
  Supercritical & $\rho$ & $T$ \\
  Ice-Liquid & $S_{ice}$ & $T$ \\
  Ice-Gas & $S_{ice}$ & $T$ \\
  Gas-Liquid & $S_{gas}$ & $T$ \\
  Ice-Gas-Liquid & $S_{ice}$ & $S_{gas}$ 
\end{tabular}
\end{center}
Solving the balance laws with this caveat requires the incorpoeration
of a statemachine. Representing the material with the known
representations thus requires two variables and an additional state
tag. For the complete states of water, the states and their possible
transitions is shown in Figure \ref{fig:statemachine}.
  \begin{figure}
    \centering
    \includegraphics[width=4in]{../slides/phase_4.png}
    \caption{\label{fig:statemachine}States and possible transitions for water.}
\end{figure}
Formulate balance equations as
\begin{equation}
R(X ; phase) = 0
\end{equation}
Solve the differentiable part:
\[\frac{\partial R}{\partial X}\Delta X^{k+1/2} = -R(X^k ; phase^k)\]
and iterate on the nondifferential part:
\[phase^{k+1},X^{k+1} = statemachine(X^{k+1/2} ; phase^k)\]
The phase switching logic has the form of a switch case illustrated in
Algorithm \ref{alg:statemachine}.

\begin{algorithm}
  \caption{\label{alg:statemachine}Statemachine transition logic}
\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{switch}\NormalTok{ phase_old:}
  \ControlFlowTok{case}\NormalTok{ gas:}
    \ControlFlowTok{if}\NormalTok{ p,T crossing boundary:}
\NormalTok{      phase_new = liquid_gas}
  \ControlFlowTok{case}\NormalTok{ liquid:}
    \ControlFlowTok{if}\NormalTok{ p,T crossing boundary:}
\NormalTok{      phase_new = liquid_gas}
  \ControlFlowTok{case}\NormalTok{ liquid_gas:}
    \ControlFlowTok{if}\NormalTok{ S_gas >= }\DecValTok{1}\NormalTok{:}
\NormalTok{      phase_new = gas}
    \ControlFlowTok{if}\NormalTok{ S_liquid >= }\DecValTok{1}\NormalTok{:}
\NormalTok{      phase_new = gas}
\end{Highlighting}
\end{Shaded}
\end{algorithm}

@@

We instead replace our problem with two balance laws and one unknown
constraint:
\begin{align}
\text{Solve for}\, \rho(t), \, h(t), \, p(t),\, \text{and}\, T(t)\,
  \text{satisfying:} \nonumber
                       \\
\partial_t \rho &  = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t (\rho h-p) & = \mathbf{k}_T(T_\infty-T)  + s + h\left(\mathbf{k}_p(p_\infty - p) + r\right) \\
\text{such that they lie on the material EOS,}\nonumber
                                                \\
  \rho,h,p,T & \in \{\text{EOS}\}\\
  \text{given }\rho,h,p,T \text{ at } t=0. \nonumber
\end{align}
The \emph{ground truth} is the experimental data in the first place; the
branching curve fit is only one realization of representing the data. We
have demonstrated that it is not possible to represent any two of these
in terms of the other two; and even if so, it is not necessarily a good
and robust fit. What if we search for a new fit?

@@

\begin{align}
\text{Solve for}\, x(t), \, y(t), \, f(t) \, \text{satisfying:} \nonumber \\
m \ddot{x} & = f x/\sqrt{x^2+y^2} \\
m \ddot{y} & = f y/\sqrt{x^2+y^2} - m g \\
  x^2 + y^2 & = R^2
  \text{given } x,y, \dot{x},\dot{y} \text{ at } t=0. \nonumber
\end{align}
Note that we are considering the nonlinear pendulum.

There are many ways to solve this canonical equation: Lagrange
multipliers (\(f(t)\) is the multiplier in the way we have written it
above), penalties, change of variables, etc. The change of variables
method exploits the Lagrangian mechanics formulation, where we state
the Lagrangian,
\begin{equation}
\mathcal{L} = \frac{1}{2}m\left(\dot{x}^2 + \dot{y}^2\right) - m g y
\end{equation}
with the constraint that \(x\) and \(y\) lie on the path.
We can introduce a new variable \(\theta\) that parameterizes
\(x=R\cos\theta\) and \(y=R\sin\theta\) and rederive the equations of
motion with the familiar,
\begin{align}
\text{Solve for}\, \theta(t) \ \text{satisfying:} \nonumber \\
\frac{\mathrm{d}}{\mathrm{d}t} \frac{\partial \mathcal{L}}{\partial
  \dot{\theta}} =
  \frac{\partial \mathcal{L}}{\partial \theta}
   \text{given } \theta,\dot{\theta} \text{ at } t=0. \nonumber
\end{align}
In this example, this lets us collapse the problem into a
single-component second-roder differential equation.

In our methodology, we do not have an explicit equation for the
constraint, but a dataset of $X$ of $(x,y)$ pairs that are on the manifold. For
our proof-of-concept test, we manufacture this data for the circular
pendulum to verify against the analytical solution:
\begin{equation}
y = \left\{ (R\cos\theta_i, R\sin\theta_i)\; | \;\theta_i \in [-5\pi/4,\pi/4] \right\}
\end{equation}
We also try a ``goofier'' looking curve with no known closed form
solution to exercise the methodology.
For the equation of state problem, this is analogous to the dataset
of $p,T,\rho,h$ 4-tuples that make the plot in Figure \ref{fig:pTrho}.

@@

We'll use the autoencoder to generate the parameterization of \(x(q)\)
and \(y(q)\). To do this, we optimize the loss function $L$ (a
distinct symbol from the Lagrangian $\mathcal{L}$),
\begin{equation}
L(x) = \left\| x-D(E(x)) \right\|_2^2
\end{equation}

\begin{figure}
  \center
\includegraphics[width=3in]{../figures/autoencoder_balance.png}
  \caption{\label{fig:balance}Illustration of the insert of the encoder to form the
  description of the system of equation. The equations
  of motion are generated in a declarative programming methodology,
  where the trained decoder can be extracted and inserted into the abstract
  syntax tree, which can then be differentiated and inserted into a
  time stepping loop.}
\end{figure}

The process of learning and generating the code is illustrated in
Figure \ref{fig:balance}.
When we want to solve the dynamics, we can treat the autoencoder the
same as \(x(\theta)\), plugging it into the equation for the Lagrangian,
\begin{equation}
  \mathcal{L}(x(q),v(q,\dot{q}))
\end{equation}
Taking the partial derivatives with respect to the feature variable and
its rate, the components to the equation of motion are build the DAE we know from physics:
\begin{equation}
  \frac{\mathrm{d}}{\mathrm{d}t}\frac{\partial \mathcal{L}}{\partial \dot{q}} =
  \frac{\partial \mathcal{L}}{\partial q}.
\end{equation}
By the chain rule, the velocities are related to the
feature space by
\begin{equation}
\dot{x} = \frac{\mathrm{d}x(q)}{\mathrm{d}t} = \frac{\partial x}{\partial q}\frac{\mathrm{d}q}{\mathrm{d}t}
\end{equation}
and similarly for the \(y\) direction.
Then plug it into a Runge Kutta (trapezoidal) to solve with Newton's
method:
\begin{equation}
\frac{\partial \mathcal{L}}{\partial \dot{q}}_i - \Delta t
\frac{\partial \mathcal{L}}{\partial q}_i = \frac{\partial
  \mathcal{L}}{\partial \dot{q}}_0 + \Delta t \frac{\partial
  \mathcal{L}}{\partial q}_0
\end{equation}

Solutions are shown in Figure \ref{fig:pendulum}.
\begin{figure}
\centering
\includegraphics[width=5in]{../figures/pendulum_q.png}
\caption{\label{fig:pendulum}Solution of the pendulum on an arbitrary latent space.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/classifier_network.png}
\caption{\label{fig:classifyingnetwork}The architecture of the classifier network designed for
  unsupervised labeling of phases.}
\end{figure}

@@

\subsection{Training Behavior}

I need to show A) learning rate graphs of all of them
B) describe the algorithm better. How many epochs?

C) How do I quantify how well it classified?

One of the surfaces colored by class is shown in
\ref{fig:classified}. 
Because the identity of the phases,XXXXXXXX
For comparison purposes, we group together liquid-gas-supercritical as one contiguous label.

\begin{figure}
  \centering
  \includegraphics[width=3in]{../slides/classifier_good.jpg}
  \caption{\label{fig:classified}Result of the unsupervised phase
    classifier. Compare this plot to the original labeled dataset
    Figure \ref{fig:pTrho}. Note the axes are in the
    nondimensionalized scaling.}
\end{figure}

@@

\begin{algorithm}
  \caption{Psuedo code of the simulation}
\begin{verbatim}
specify EOS and a which network
load network from database
build DAE graph
input problem properties and initial value
solve for q0: D(q0) = s0
loop t=0 to t_max:
    solve for q[i]: R(q[i],q[i-1])=0
    decode s[i] = D(q[i])
\end{verbatim}
  \end{algorithm}
